Group Theory in Physics (Course Note)

### 2016-11-14 Lecture note - Group Theory in Physics/sections/20161107.tex
Almost finised the Irreps if D_n and C_n. (representation, group theory)
Finished all proof of recent classes' contents, just need to add my own scanned notes.

\subsection{Finding Representation}
\label{sec:Finding-Representation}
One important work in group theory is to find all its irreducible
representations. The book \cite{book} suggests we first find the
characters of irreducible representations, and list them in a
character table, and then find the matrix of those irreducible
representations.

@@

\begin{fact}
    Also, a the direct product of two irreducible representations is also
    a representation. If the dimension of one of the representation is
    $1$, then their product is still a irreducible representation.
\end{fact}
\subsection{Self-conjugate representation and Real Representation}
\label{sec:Self-conjugate-representation-and-Real-Representation}

It's easy to see that for any representation $D$, conjugating ever
matrix $D(g)^*$ will give another representation. We define
\begin{defi}[Conjugate Representation]
\nomenclature{Conjugate Representation}{\nomrefpage.}
    The conjugate representation of a given representation $D$ is obtained
    by complex conjugate all its matrices $D(g)^*,\,g\in G$.
\end{defi}
\begin{defi}[Self-conjugate Representation]
\nomenclature{Self-conjugate Representation}{\nomrefpage.}
    A representation $D$ is called self-conjugate if and only if it is
    equivalent (but not necessarily equal) to its conjugate
    representation $D^*$.
\end{defi}
\begin{fact}
    The character of a self-conjugate representation is a real
    function. Since (assume $XD(g)X^{-1}=D^*(g)$):
    $$
        \Tr(XD(g)X^{-1}) = \Tr(X^{-1}XD(g)) = \Tr(D(g)) = \Tr(D^*(g))
        = \Tr(D(g))^*
    $$
\end{fact}

@@

\subsection{Subduced Representation}
\label{sec:Subduced-Representation}
\marginpar{To be finished.}
Let $G$ be a group. Let $\mathcal{C}_a$ denotes it conjugacy classes.
Let $D^j$ be its irreducible representations, with dimension $m_j$,
character $\chi^j_a$. Assume it has a subgroup $H=\{T_1=E,
T_2,\cdots,T_h\}$, where $h=|H|$. Let $n=|G|/h$ be its index. Denotes
its coset by $R_rH$. With $R_1:=E$, any element in $G$ can be denotes
as $R_r T_t$ (assume $R_r$ is chosen). Let $\bar{\mathcal{C}}_\beta$
denotes conjugacy classes of the subgroup $H$, with character
$\bar{\chi}^k_\beta$, dimension $\bar{m}_k$.

>>>>

\subsection{Finding Representation}
\label{sec:Finding-Representation}
% TODO Polish the arragement of this section.
One important work in group theory is to find all its irreducible
representations. The book \cite{book} suggests we first find the
characters of irreducible representations, and list them in a
character table, and then find the matrix of those irreducible
representations.

@@

\begin{fact}
    Also, a the direct product of two irreducible representations is also
    a representation. If the dimension of one of the representation is
    $1$, then their product is still a irreducible representation.
\end{fact}
\begin{fact}
    \label{fact:character-of-identity}
    Another helpful point is that, for all irreducible
    representations, the identity element is always mapped to the
    identity matrix. So the character for this unit class is just the
    dimension of the representation.
\end{fact}
\subsection{Self-conjugate representation and Real Representation}
\label{sec:Self-conjugate-representation-and-Real-Representation}

It's easy to see that for any representation $D$, conjugating ever
matrix $D(g)^*$ will give another representation. We define
\begin{defi}[Conjugate Representation]
\nomenclature{Conjugate Representation}{\nomrefpage.}
    The conjugate representation of a given representation $D$ is obtained
    by complex conjugate all its matrices $D(g)^*,\,g\in G$.
\end{defi}
\begin{defi}[Self-conjugate Representation]
\nomenclature{Self-conjugate Representation}{\nomrefpage.}
    A representation $D$ is called self-conjugate if and only if it is
    equivalent (but not necessarily equal) to its conjugate
    representation $D^*$.
\end{defi}
Note that the character of a self-conjugate representation is a real
function. Since (assume $XD(g)X^{-1}=D^*(g)$):
$$
    \Tr(XD(g)X^{-1}) = \Tr(X^{-1}XD(g)) = \Tr(D(g)) = \Tr(D^*(g))
    = \Tr(D(g))^*
$$
In fact, we have
\begin{thm}
    A represents is self-conjugate if and only if its character is
    real-valued.
\end{thm}
\begin{proof}
    If it is self-conjugate, clearly it is real-valued. On the other
    hand, if the character is real-valued, then by
    \begin{equation}
        \chi_{D^*(g)} = \Tr(D^*(g)) = \Tr(D(g))^* = \chi^*_{D(g)}
        = \chi_{D(g)}
    \end{equation}
    The conjugate representation has the same character, hence is
    equivalent to original representation by theorem \ref{thm:20161031-character-determine-rep}
\end{proof}

Fun fact. Since all representation of finite group are equivalent to a
unitary one, if for en element $g\in G$ its character $\chi(g)$ is
complex, then its inverse $g^{-1}$ has character $\chi(g)^*$.
Therefore, for a representation which is not self-conjugate, its
character will has complex value existing in pairs, unless $g=g^{-1}$.

@@

\subsection{Subduced Representation}
\label{sec:Subduced-Representation}
\marginpar{To be finished.} %TODO
Let $G$ be a group. Let $\mathcal{C}_a$ denotes it conjugacy classes.
Let $D^j$ be its irreducible representations, with dimension $m_j$,
character $\chi^j_a$. Assume it has a subgroup $H=\{T_1=E,
T_2,\cdots,T_h\}$, where $h=|H|$. Let $n=|G|/h$ be its index. Denotes
its coset by $R_rH$. With $R_1:=E$, any element in $G$ can be denotes
as $R_r T_t$ (assume $R_r$ is chosen). Let $\bar{\mathcal{C}}_\beta$
denotes conjugacy classes of the subgroup $H$, with character
$\bar{\chi}^k_\beta$, dimension $\bar{m}_k$.

### 2016-11-14 Lecture note - Group Theory in Physics/sections/point-group.tex
Finished all proof of recent classes' contents, just need to add my own scanned notes.

But we can have another more mathematical definition of Dihedral
group:
\marginpar{To be added}
\paragraph{Conjugacy Classes}
There is some important difference between the case when $n$ is odd
and when $n$ is even. When $n$ is odd, the $2$-fold axis are all
linked with the vortices of the $n$-polygon. On the contrary, when $n$
is even, there is two different $2$-fold axis. The first $n/2$ axises
are those linked with the vortices, the second $n/2$ axises are those
cut through the opposite edges.

Note that by fact \ref{fact:20161010-srs} one can see that such
difference will note make the two classes in even $n$ the same
conjugacy class. On the contrary, by the same fact
\ref{fact:20161010-srs} one sees that those $2$-fold axises are in the
same conjugacy class when $n$ is odd.

Also denotes one $2\pi/n$ rotation around the $N$-fold axis by $T$,
then by this very same fact \ref{fact:20161010-srs}, $T$ and $T^{-1}$
lies in the same conjugacy class, related by a reflection around a
$2$-fold axis. This is obvious a self-conjugate class. (Note that,
when only $2\pi/n$ rotation is considered, they form an abelian group
and each rotation forms a conjugacy class.) Similarly for the
conjugacy class $\{T^m, T^{-m}\}$, where $m$ is an integer.

In summary, this discussion shows that the self-conjugate classes in $D_n$
are:
\begin{fact}
    When $n$ is odd (Let $n=2n'+1$): The identity class, the class of
    all $2$-fold axis. The class of $\{ T^m, T^{-m}\}$, where
    $m=1,2,\cdots, n'$. (Why $m\leq n'$? Think about the case when
    $n=5,n'=2$.). In total $n'+2=\frac{n+3}{2}$ conjugacy
    classes/irreducible representations. All conjugacy classes are
    self-conjugate.
\end{fact}
\begin{fact}
    When $n$ is even (Let $n=2n'$): The identity class, the two
    classes of all $2$-fold axis, as discussed above. The class of
    $\{ T^m, T^{-m}\}$, where $m=1,2,\cdots, n'$. In total
    $n'+3=\frac{n+6}{2}$. All conjugacy classes are self-conjugate.
\end{fact}

@@

@@

There is two $1$-dimensional irreducible representations, formed using
the invariant subgroup $C_{2n'+1}$. Since $4n'+2 = 2^2 n'+2$, one
guesses that the remaining ones are $n$ $2$-dimensional irreducible
representation. They are constructed (construction process could be
found in \cite{book}) to be (notice $D_{2n'+1}$ is generated by
$T_{2n'+1}$, the rotation $2\pi/(2n'+1)$, and $C'_2$, any one of the
$2$-fold reflection):
\begin{align}
    \bar{D}^{E_j}(C_{2n'+1}) &=\left( \begin{array}{cc}
    \cos(\frac{2j\pi}{2n'+1}) & -\sin(\frac{2j\pi}{2n'+1}) \\
    \sin(\frac{2j\pi}{2n'+1}) & \cos(\frac{2j\pi}{2n'+1}) \\
        \end{array} \right) \\
    \bar{D}^{E_j}(C'_{2}) &= \left( \begin{array}{cc}
                 1 & 0 \\
                 0 & -1 \\ 
                 \end{array} \right)
\end{align}
where $j=1,2,\cdots,n$.

The case for even $n$ is simialr (let $n'=n/2$). Except that here one
as three invariant subgroup of index $2$ (see dicusson about invariant
subgroup above). So it has three $1$-dimensional irreducible
representations, and $n-1$ $2$-dimensional irreducible representations.

When $n$ is odd, let $n=2n'$. (TBD)

%TODO Finish it.

>>>>

But we can have another more mathematical definition of Dihedral
group. To introduce that definition, let me first explain some facts
and discuss a little bit about its conjugacy classes.

There is some important difference between the case when $n$ is odd
and when $n$ is even. When $n$ is odd, the $2$-fold axis are all
linked with the vortices of the $n$-polygon. On the contrary, when $n$
is even, there is two different $2$-fold axis. The first $n/2$
$2$-fold axises are those linked with the vortices, the second $n/2$
$2$-fold axises are those cutting through the opposite edges.

\paragraph{Conjugacy Classes}
Note that by fact \ref{fact:20161010-srs} one can see that such
difference mentioned will make the two classes of $2$-fold axises in
even $n$ each form their own conjugacy class. On the contrary, by the
same fact \ref{fact:20161010-srs} one sees that when $n$ is odd, all
the $2$-fold axises are in the same conjugacy class. Note that a
reflection is its own inverse, so in either cases ($n$ odd or even),
we have found self-conjugate classes.

The next conjugacy classes are about rotation. Denotes one $2\pi/n$
rotation around the $N$-fold axis by $T$, then by this very same fact
\ref{fact:20161010-srs}, $T$ and $T^{-1}$ lies in the same conjugacy
class, related by a reflection around a $2$-fold axis. This is obvious
a self-conjugate class. (Note that, when only $2\pi/n$ rotation is
considered, they form an abelian group and each rotation forms a
conjugacy class.) Similarly for the conjugacy class $\{T^m, T^{-m}\}$,
where $m$ is an integer.

In summary, this discussion shows that the conjugacy classes (all are
self-conjugate) in $D_n$
are:
\begin{fact}
    When $n$ is odd (Let $n=2n'+1$): The identity class, the class of
    all $2$-fold axis. The class of $\{ T^m, T^{-m}\}$, where
    $m=1,2,\cdots, n'$. (Why $m\leq n'$? Think about the case when
    $n=5,n'=2$.). In total $n'+2=\frac{n+3}{2}$ conjugacy
    classes/irreducible representations. All conjugacy classes are
    self-conjugate.
\end{fact}
\begin{fact}
    When $n$ is even (Let $n=2n'$): The identity class, the two
    classes of all $2$-fold axis, as discussed above. The class of
    $\{ T^m, T^{-m}\}$, where $m=1,2,\cdots, n'$. In total
    $n'+3=\frac{n+6}{2}$. All conjugacy classes are self-conjugate.
\end{fact}

\paragraph{Formal definition or $D_n$} Having discussed about
conjugacy classes, one sees that $D_n$ can be defined by generators.
Clearly, when $n$ is odd, $D_n$ is generated by one rotation $2\pi/n$
and one reflection (denoted $C'_2$ in the book \cite{book}). When $n$
is even, one might thinks that we have three generators, i.e. the
rotation by $2\pi/n$, one reflection about vortex $C'_2$, and one
reflection about middle of edges $C''_2$. But actually it only needs
two. The reflection about middle of edges can be produced by the
rotation and the reflection about vortex, or vice versa. See this
note for example:

% TODO Add scanned notes.

So all in all, $D_n$ can be generated by just two elements. They are:
\begin{fact}
\label{fact:point-group-formal-def-dn}
\begin{align}
    D_n &\equiv \braket{R,S|R^n=1,S^2=1,SRS^{-1}=R^{-1}} \\
        &\equiv \braket{R,S|R^n=1,S^2=1,(SR)^2=1}
\end{align}
Here the first line symbolizes that rotation $R$ and $R^{-1}$ are in
the same conjugacy class. The second is more beautiful and symbolizes
that a roto-reflection is idempotent.
\end{fact}

@@

I present only vague discussion below. For details, please to ...
actually I don't know which book. The textbook \cite{book} is too
vague too. Perhaps I should find some more detailed books.
% TODO Add better books.

@@

There is two $1$-dimensional irreducible representations, formed using
the invariant subgroup $C_{2n'+1}$. Since $4n'+2 = 2^2 n'+2$, one
guesses that the remaining ones are $n$ $2$-dimensional irreducible
representation. They are constructed (construction process could be
found in \cite{book}) to be (notice $D_{2n'+1}$ is generated by
$T_{2n'+1}$, the rotation $2\pi/(2n'+1)$, and $C'_2$, any one of the
$2$-fold reflection):
\begin{align}
    \bar{D}^{E_j}(C_{n}) &=\left( \begin{array}{cc}
    \cos(\frac{2j\pi}{n}) & -\sin(\frac{2j\pi}{n}) \\
    \sin(\frac{2j\pi}{n}) & \cos(\frac{2j\pi}{n}) \\
        \end{array} \right) \\
    \bar{D}^{E_j}(C'_{2}) &= \left( \begin{array}{cc}
                 1 & 0 \\
                 0 & -1 \\ 
                 \end{array} \right)
\end{align}
where $j=1,2,\cdots,n$.

The case for even $n$ is simialr (let $n'=n/2$). Except that now it
has three invariant subgroup of index $2$ (see dicusson about invariant
subgroup above). So it has $2+2+2-2=4$ $1$-dimensional irreducible
representations ($-2$ for the repeated identity representation), and
$n'-1$ $2$-dimensional irreducible representations.
(we have $\sum_{j=1}^{n'+3}m_j^2 = 4n' = (n-1)2^2 + 4\cdot 1$.)
The result is exactly the same as in the the add $n$ case, and this is
why I did not write $2n'+1$ but use $n$ in the above equations.
